---
title: "Principal Component Analysis"
author: "David Neuzerling"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides:
      ratio: "16:9"
---

```{r setup, include=FALSE}
library(tidyverse)
options(htmltools.dir.version = FALSE)
```

class: center, middle

# xaringan

### /ʃaː.'riŋ.ɡan/

Not PowerPoint

---

# Intuitive definition

> PCA can be thought of as fitting a $p$-dimensional ellipsoid to the data, where each axis of the ellipsoid represents a principal component.
> 
> - Wikipedia

---
class: middle, center

<blockquote class="twitter-tweet"><p lang="en" dir="ltr">doing math:<br>step 1: relief when your problem reduces to linear algebra <br>step 2: panic when you realize you somehow don&#39;t actually know any linear algebra</p>&mdash; Aleksandra Sobieska (@combinatola) <a href="https://twitter.com/combinatola/status/997147577173925888?ref_src=twsrc%5Etfw">May 17, 2018</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script> 

---
# Linear algebra you forgot in 1st-year uni

--

$$
\left(\begin{array}{cc} 
2 & 3 \\\\
2 & 1
\end{array}\right) \times
\left(\begin{array}{c} 
6 \\\\
4
\end{array}\right) = 
\left(\begin{array}{c} 
24 \\\\
16
\end{array}\right) = 
4 \left(\begin{array}{c} 
6 \\\\
4
\end{array}\right)
$$ 

--

$(6, 4)^T$ is an **eigenvector** of the square matrix, and 4 is its **eigenvalue**.

--

The square matrix makes the eigenvector *bigger*, but it doesn't change its *direction*.

---
# Linear algebra you forgot in 1st-year uni

* Eigenvectors of real symmetric matrices are real.
* Eigenvectors of real symmetric matrices are orthogonal to each other.
* For an $n \times n$ real symmetric matrix, there are $n$ eigenvectors.

---
# Linking it back to statistics

Covariance matrices are real and symmetric.

```{r diamonds-covariance-matrix}
diamonds %>% select_if(is.numeric) %>% cov %>% round(1)
```

---
# Show me

```{r square-plot, include = FALSE}
square_ggplot <- function(data, pc_labels = FALSE) {
  p <- ggplot(data, aes(x = wt, y = mpg)) + 
  geom_point() + 
  xlim(c(-2.5, 2.5)) +
  ylim(c(-2.5, 2.5)) +
  coord_fixed() + # makes the plot square
  theme(text = element_text(size = 20))
  
  if (pc_labels) {
    p <- p + xlab("PC1") + ylab("PC2")
  }
  
  return(p)
}
```

.pull-left[
```{r mtcars-scaled}
mtcars_scaled <- mtcars %>% 
  select(wt, mpg) %>% 
  mutate_all(scale)
```

```{r mtcars-scaled-plot, eval = FALSE}
# square_ggplot is a helper function
# that uses ggplot2
mtcars_scaled %>% square_ggplot
```
]

.pull-right[
```{r mtcars-scaled-plot-out, ref.label = "mtcars-scaled-plot", echo = FALSE}
```
]

---
# Eigenstuff

```{r mtcars-prcomp}
eigenstuff <- mtcars_scaled %>% select(wt, mpg) %>% cor %>% eigen
eigenstuff
```

--

```{r extracting-pc-values}
eigenvector_1 <- eigenstuff$vectors[,1] %>% {c("wt" = .[1], "mpg" = .[2])}
eigenvalue_1 <- eigenstuff$values[1]

eigenvector_2 <- eigenstuff$vectors[,2] %>% {c("wt" = .[1], "mpg" = .[2])}
eigenvalue_2 <- eigenstuff$values[2]
```

---
# Show me

```{r geom-arrow, include = FALSE}
arrow <- function(x_start = 0, x_end, y_start = 0, y_end, scale = 1) {
  geom_segment(
    aes(
      x = x_start,
      xend = x_end * scale,
      y = y_start,
      yend = y_end * scale
    ),
    arrow = ggplot2::arrow(length = unit(0.5, "cm")),
    col = "red"
  )
}
```

.pull-left[
```{r mt-cars-scaled-with-vectors, eval = FALSE}
# `arrow` is a helper function
mtcars_scaled %>% square_ggplot +
  arrow(x_end = eigenvector_1[["wt"]],
        y_end = eigenvector_1[["mpg"]]) +
  arrow(x_end = eigenvector_2[["wt"]],
        y_end = eigenvector_2[["mpg"]])
```
]

.pull-right[
```{r mt-cars-scaled-with-vectors-out, ref.label = "mt-cars-scaled-with-vectors", echo = FALSE}
```
]

---

# Scaling with eigenvalues

.pull-left[
```{r mt-cars-scaled-with-vectors-and-values, eval = FALSE}
mtcars_scaled %>% square_ggplot +
  arrow(x_end = eigenvector_1[["wt"]],
        y_end = eigenvector_1[["mpg"]],
        scale = eigenvalue_1) +
  arrow(x_end = eigenvector_2[["wt"]],
        y_end = eigenvector_2[["mpg"]],
        scale = eigenvalue_2)
```
]

.pull-right[
```{r mt-cars-scaled-with-vectors-and-eigenvalues-out, ref.label = "mt-cars-scaled-with-vectors-and-values", echo = FALSE}
```
]

---
# PCA steps

1. Normalise your data
1. Transform with eigenvectors
1. Drop _some_ eigenvectors with the lowest eigenvalues
1. Do stuff
1. Transform back

---

# Original data

.pull-left[

```{r original-data, eval = FALSE}
mtcars_scaled %>% 
  square_ggplot
```
]

.pull-right[
```{r original-data-out, ref.label = "original-data", echo = FALSE, warning = FALSE}
```
]

---

# Change of basis

.pull-left[

\begin{align*}
\text{data} \;& \cdot \\
\text{eigenvector matrix} \;&
\end{align*}

```{r change-of-basis, eval = FALSE}
as.matrix(mtcars_scaled) %*% # dot product
  eigenstuff$vectors %>% 
  data.frame %>% 
  setNames(c("wt", "mpg")) %>% 
  square_ggplot(pc_labels = TRUE)
```
]

.pull-right[
```{r change-of-basis-out, ref.label = "change-of-basis", echo = FALSE, warning = FALSE}
```
]

---

# Dropping a dimension

.pull-left[

\begin{align*}
\text{data} \;& \cdot \\
\text{eigenvector matrix} \;&
\end{align*}

```{r change-of-basis-with-residuals, eval = FALSE}
as.matrix(mtcars_scaled) %*%
  eigenstuff$vectors %>% 
  data.frame %>% 
  setNames(c("wt", "mpg")) %>% 
  square_ggplot(pc_labels = TRUE) +
  geom_segment(aes(x = wt, y = mpg,
                   xend = wt, yend = 0),
               colour = "blue")
```

```{r eigenvalue-reminder}
eigenvalue_1
eigenvalue_2 # The smallest, so we drop it
```
]

.pull-right[
```{r change-of-basis-with-residuals-out, ref.label = "change-of-basis-with-residuals", echo = FALSE, warning = FALSE}
```
]

---

# Dropping a dimension

.pull-left[
\begin{align*}
\text{data} \;& \cdot \\
\text{eigenvector matrix} \;& \cdot \\
\text{projection matrix} \:&
\end{align*}

```{r drop-dimension, eval = FALSE}
as.matrix(mtcars_scaled) %*%
  eigenstuff$vectors %*%
  rbind(c(1, 0),
        c(0, 0)) %>% 
  data.frame %>% 
  setNames(c("wt", "mpg")) %>% 
  square_ggplot(pc_labels = TRUE)
```
]

.pull-right[
```{r drop-dimension-out, ref.label = "drop-dimension", echo = FALSE, warning = FALSE}
```
]

---

# Transforming back

.pull-left[
\begin{align*}
\text{data} \;& \cdot \\
\text{eigenvector matrix} \;& \cdot \\
\text{projection matrix} \:& \cdot \\
\text{inverse eigenvector matrix} \;&
\end{align*}

```{r transforming-back, eval = FALSE}
as.matrix(mtcars_scaled) %*%
  eigenstuff$vectors %*%
  rbind(c(1, 0),
        c(0, 0))  %*%
  solve(eigenstuff$vectors) %>% 
  data.frame %>% 
  setNames(c("wt", "mpg")) %>% 
  square_ggplot
```
]

.pull-right[
```{r transforming-back-out, ref.label = "transforming-back", echo = FALSE, warning = FALSE}
```
]
